A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-03-19 graph_syntax.tex
typo
changed all instances of 'free' to explicit

MDAO problems require that information be passed between sets of analyses. When information from the output of one analysis 
block is passed, or connected, to the input of another analysis block, a new free edge is added 
connecting the two variable nodes involved in the exchange. The edge is free because, unlike the edges 
within an analysis block, it could be added or removed depending on the demands of the specific problem.

@@

Objective and constraints are also constructs that need to be identified by engineers 
as part of the design process. In the case of objective functions single output values 
could be selected, but commonly multiple values are assembeled together via simple 
athrimetic opperations to form the final objective function. For example, the problem from 
Fig. \ref{f:design vars} might have an objective function of $z^2$ or $2y + z$. Constraints are 
always given in the form of either an inequality of an equality, e.g. 
$\frac{2y}{10}-1<0$ or $3x+2y - 5=0$. In both cases, an engineer explictly identifies variables 
or groups of variables and assigns an simple, explicit relationship between them.

For the proposed syntax objectives and constraints are represented through the use of 
expression nodes. These nodes contain the actual mathematical expression used to define the 
objective or constraint value. Objective or constraint nodes are distinguished
via the presence of a simple expression (objective) vs an inequality/equality (constraint). 
Fig. [insert ref] shows the notional problem graph augmented with objective and 
constraint values.

@@

\begin{comment}
The MDO problem formulation concepts represented by nodes are given in 
Table \ref{t:node representation}, and the concepts represented by edges are 
given in Table \ref{t:edge representation}.
% \begin{itemize}
% \item An objective or constraint is indicated by a variable node with no outgoing edges and with the incoming edges being directed from only other variable nodes.
% \item A global input is a variable node with no edges directed in and with at least two edges directed out to different variable nodes.
% \item A design variable is represented by a variable node with no edges directed in and on fixed edge directed out.
% \end{itemize}
\begin{table}[h!]
 \begin{center}
  \caption{Problem formulation concept represented by a node}
  \label{t:node representation}
  \begin{tabular}{ccc} \hline 
edges directed in & edges directed out & node representation \\ \hline
only free edges & none & objective or constraint\\
none & at least two free edges & global input \\
none & one fixed edge & design variable \\ \hline
  \end{tabular}
 \end{center}
 \vspace{-15pt}
\end{table}
\begin{table}[h!]
 \begin{center}
  \caption{Problem formulation concept represented by an edge}
  \label{t:edge representation}
  \begin{tabular}{ccc} \hline 
from node & to node & edge representation \\ \hline
variable & variable & connection/passing of a variable\\
variable & model & local input \\
model & variable & local output \\
model & model & function call \\ \hline
  \end{tabular}
 \end{center}
 \vspace{-15pt}
\end{table}
%`Connection' edges connect the local outputs from one analysis block to input nodes.
\end{comment}
\subsection{Local vs Global Properties}

@@

\subsection{Maximal Connectivity Graph}
To construct the maximal connectivity graph, we assume that a set of codes, global inputs, and objectives and constraints (collectively called global outputs). The codes are represented by analysis blocks $A_i=(V_{A_i},E_{A_i}), \ i\in \{1,2,\ldots,m\}=I$, the global inputs are represented a set of variable nodes $V_\txt{in}$, and the global outputs are represented by a set of variable nodes $V_\txt{out}$. We assume that $V_\txt{in}$, $V_\txt{out}$, and $A_i$ are given, and that any potential connection between variables is given in the form of the free edges in the set $C_M$. 
Then we may construct the maximal connectivity graph $M=(V_M,E_M)$ as
\begin{IEEEeqnarray*}{rCl}
V_M & = & V_\txt{in} \cup V_\txt{out} \cup \left( \bigcup_{i \in I} V_{A_i} \right), \\
E_M & = & C_M \cup \left( \bigcup_{i \in I} E_{A_i} \right),
\end{IEEEeqnarray*}
The MCG $M$ is uniquely determined by the given set of analysis blocks, the required outputs, and the given global inputs. In the cases where the set of global inputs $I$ is not known a priori, the process of obtaining the FPF will reveal the required inputs, as discussed subsequently.

@@

The set $C_F$ is a set containing free edges representing connections. Since $C_M$ is the set of all potential connections, we must have $C_F \subset C_M$. While no other conditions are explicitly stated for $C_F$, requirement (6) is actually a requirement on both $I$ and $C_F$.

\subsection{Obtaining the Fundamental Problem Formulation Graph}
In general, there may be multiple different graphs that satisfy the FPF conditions, though there may be none at all. Here, we describe a process for obtaining an FPF by starting with the MCG and disconnecting free edges until the FPF conditions are met. Then the problem is reduced to deciding which free edges to remove.

@@

\begin{enumerate}
\item The first step is to detect holes and disconnect the free edges following them. These free edges are removed because they represent variables which cannot be determined because the analysis function does not have adequate inputs. The set of variable nodes which are holes is created as
\begin{equation}
H = \{v \in V \st t_\txt{node}(v) = \txt{`variable'} \txt{ and } \txt{deg}^-(v) < \txt{deg}_l^-(v) \},
\end{equation}
which is the set of variable nodes with few incoming edges than are allowed by the lower indegree limit.
Then the updated set of edges is created by removing the first set of free edges following the hole:
\begin{equation}
C_{F,1} = C_{F,0} \setminus \{(x,y) \in E \st x=E^3(v) \txt{ for } v \in H\},
\end{equation}
%deg-v is now recaculated for the new C. 
and this step is demonstrated by Fig.~\ref{f:hole}. Because removing these edges can create new holes, this step must be repeated until no more holes are found.
\begin{figure}[htb!]
	\begin{center}
	\includegraphics[width=3in]{images/analysis_block_hole}
	\end{center}
	\vspace{-20pt}
\caption{Example variable node indicating a hole.}
\label{f:hole}
\end{figure}

\item The second step is to detect collisions and to then disconnect enough free edges so that the collisions are resolved. The set of variable nodes which contain collisions is created as
\begin{equation}
S_\txt{nodes} = \{v \in V \st t_\txt{node}(v) = \txt{`variable'} \txt{ and } \txt{deg}^-(v) > \txt{deg}_u^-(v) \}.
\end{equation}
For each collision node we can construct a set contained the edges directed in. The set containing all of these sets is constructed as
\begin{equation}
S_\txt{edges} = \big \{ \{(x,y) \in E\} \st y \in S_\txt{nodes} \big \}
\end{equation}
Let $J=\{1,2,\ldots,|S_\txt{edges}|\}$ be an indexing set for $S_\txt{edges}$ such that each $S_{\txt{edges},j}$ corresponds to a set in $S_\txt{edges}$ for $j \in J$. An examle collision is shown in Fig.~\ref{f:collision} to indicate the definition of $S_{\txt{edges},j}$. We can assume that $J$ also indexes $S_\txt{nodes}$ because there is a one--to--one correspondence between the elements in $S_\txt{nodes}$ and the elements in $S_\txt{edges}$. Then we may construct sets of edges as
\begin{equation}
B_j = \big \{e_k \in S_{\txt{edges},j} \st k \in \{1,2,\ldots,K\} \txt{ with } K \leq \txt{deg}_u^-(v_j) \big \}, \ j \in J,
\end{equation}
which means that each set $B_j$ is constructed from the set $S_{\txt{edges},j}$ by taking only as many edges as are allowed by the upper indegree limit of $v_j$. The construction of each $B_j$ corresponds to making a decision on which edges to include and which edges not to include. Then let
\begin{equation}
C_{F,2} = \{ e \in C_{F,1} \st e \in B_j \txt{ for some } j \in J\}.
\end{equation}
\begin{figure}[htb!]
	\begin{center}
	\includegraphics[width=3in]{images/analysis_block_collision}
	\end{center}
	\vspace{-20pt}
\caption{Example variable node indicating a collision.}
\label{f:collision}
\end{figure}

>>>>

MDAO problems require that information be passed between sets of analyses. When information from the output of one analysis 
block is passed, or connected, to the input of another analysis block, a new explicit edge is added 
connecting the two variable nodes involved in the exchange. Explicit edges are free because, unlike the edges 
within an analysis block, they could be added or removed depending on the demands of the specific problem. 
In other words, explicit edges are created by engineers linking the output of one tool to the input 
of another one.

@@

Objective and constraints are also constructs that need to be identified by engineers 
as part of the design process. In the case of objective functions single output values 
could be selected, but commonly multiple values are assembeled together via simple 
athrimetic opperations to form the final objective function. For example, the problem from 
Fig. \ref{f:design vars} might have an objective function of $z^2$ or $2y + z$. Constraints are 
always given in the form of either an inequality of an equality, e.g. 
$\frac{2y}{10}-1<0$ or $3x+2y - 5=0$. In both cases, an engineer explictly identifies variables 
or groups of variables and assigns a simple, explicit relationship between them.

For the proposed syntax, objectives and constraints are represented through the use of 
expression nodes. These nodes contain the actual mathematical expression used to define the 
objective or constraint value. Objective or constraint nodes are distinguished
via the presence of a simple expression (objective) vs an inequality/equality (constraint). 
Fig. [insert ref] shows the notional problem graph augmented with objective and 
constraint values.

@@

\begin{comment}
The MDO problem formulation concepts represented by nodes are given in 
Table \ref{t:node representation}, and the concepts represented by edges are 
given in Table \ref{t:edge representation}.
% \begin{itemize}
% \item An objective or constraint is indicated by a variable node with no outgoing edges and with the incoming edges being directed from only other variable nodes.
% \item A global input is a variable node with no edges directed in and with at least two edges directed out to different variable nodes.
% \item A design variable is represented by a variable node with no edges directed in and on fixed edge directed out.
% \end{itemize}
\begin{table}[h!]
 \begin{center}
  \caption{Problem formulation concept represented by a node}
  \label{t:node representation}
  \begin{tabular}{ccc} \hline 
edges directed in & edges directed out & node representation \\ \hline
only explicit edges & none & objective or constraint\\
none & at least two explicit edges & global input \\
none & one fixed edge & design variable \\ \hline
  \end{tabular}
 \end{center}
 \vspace{-15pt}
\end{table}
\begin{table}[h!]
 \begin{center}
  \caption{Problem formulation concept represented by an edge}
  \label{t:edge representation}
  \begin{tabular}{ccc} \hline 
from node & to node & edge representation \\ \hline
variable & variable & connection/passing of a variable\\
variable & model & local input \\
model & variable & local output \\
model & model & function call \\ \hline
  \end{tabular}
 \end{center}
 \vspace{-15pt}
\end{table}
%`Connection' edges connect the local outputs from one analysis block to input nodes.
\end{comment}
\subsection{Local vs Global Properties}

@@

\subsection{Maximal Connectivity Graph}
To construct the maximal connectivity graph, we assume that a set of codes, global inputs, and objectives and constraints (collectively called global outputs). The codes are represented by analysis blocks $A_i=(V_{A_i},E_{A_i}), \ i\in \{1,2,\ldots,m\}=I$, the global inputs are represented a set of variable nodes $V_\txt{in}$, and the global outputs are represented by a set of variable nodes $V_\txt{out}$. We assume that $V_\txt{in}$, $V_\txt{out}$, and $A_i$ are given, and that any potential connection between variables is given in the form of the explicit edges in the set $C_M$. 
Then we may construct the maximal connectivity graph $M=(V_M,E_M)$ as
\begin{IEEEeqnarray*}{rCl}
V_M & = & V_\txt{in} \cup V_\txt{out} \cup \left( \bigcup_{i \in I} V_{A_i} \right), \\
E_M & = & C_M \cup \left( \bigcup_{i \in I} E_{A_i} \right),
\end{IEEEeqnarray*}
The MCG $M$ is uniquely determined by the given set of analysis blocks, the required outputs, and the given global inputs. In the cases where the set of global inputs $I$ is not known a priori, the process of obtaining the FPF will reveal the required inputs, as discussed subsequently.

@@

The set $C_F$ is a set containing explicit edges representing connections. Since $C_M$ is the set of all potential connections, we must have $C_F \subset C_M$. While no other conditions are explicitly stated for $C_F$, requirement (6) is actually a requirement on both $I$ and $C_F$.

\subsection{Obtaining the Fundamental Problem Formulation Graph}
In general, there may be multiple different graphs that satisfy the FPF conditions, though there may be none at all. Here, we describe a process for obtaining an FPF by starting with the MCG and disconnecting explicit edges until the FPF conditions are met. Then the problem is reduced to deciding which explicit edges to remove.

@@

\begin{enumerate}
\item The first step is to detect holes and disconnect the explicit edges following them. These explicit edges are removed because they represent variables which cannot be determined because the analysis function does not have adequate inputs. The set of variable nodes which are holes is created as
\begin{equation}
H = \{v \in V \st t_\txt{node}(v) = \txt{`variable'} \txt{ and } \txt{deg}^-(v) < \txt{deg}_l^-(v) \},
\end{equation}
which is the set of variable nodes with few incoming edges than are allowed by the lower indegree limit.
Then the updated set of edges is created by removing the first set of explicit edges following the hole:
\begin{equation}
C_{F,1} = C_{F,0} \setminus \{(x,y) \in E \st x=E^3(v) \txt{ for } v \in H\},
\end{equation}
%deg-v is now recaculated for the new C. 
and this step is demonstrated by Fig.~\ref{f:hole}. Because removing these edges can create new holes, this step must be repeated until no more holes are found.
\begin{figure}[htb!]
	\begin{center}
	\includegraphics[width=3in]{images/analysis_block_hole}
	\end{center}
	\vspace{-20pt}
\caption{Example variable node indicating a hole.}
\label{f:hole}
\end{figure}

\item The second step is to detect collisions and to then disconnect enough explicit edges so that the collisions are resolved. The set of variable nodes which contain collisions is created as
\begin{equation}
S_\txt{nodes} = \{v \in V \st t_\txt{node}(v) = \txt{`variable'} \txt{ and } \txt{deg}^-(v) > \txt{deg}_u^-(v) \}.
\end{equation}
For each collision node we can construct a set contained the edges directed in. The set containing all of these sets is constructed as
\begin{equation}
S_\txt{edges} = \big \{ \{(x,y) \in E\} \st y \in S_\txt{nodes} \big \}
\end{equation}
Let $J=\{1,2,\ldots,|S_\txt{edges}|\}$ be an indexing set for $S_\txt{edges}$ such that each $S_{\txt{edges},j}$ corresponds to a set in $S_\txt{edges}$ for $j \in J$. An examle collision is shown in Fig.~\ref{f:collision} to indicate the definition of $S_{\txt{edges},j}$. We can assume that $J$ also indexes $S_\txt{nodes}$ because there is a one--to--one correspondence between the elements in $S_\txt{nodes}$ and the elements in $S_\txt{edges}$. Then we may construct sets of edges as
\begin{equation}
B_j = \big \{e_k \in S_{\txt{edges},j} \st k \in \{1,2,\ldots,K\} \txt{ with } K \leq \txt{deg}_u^-(v_j) \big \}, \ j \in J,
\end{equation}
which means that each set $B_j$ is constructed from the set $S_{\txt{edges},j}$ by taking only as many edges as are allowed by the upper indegree limit of $v_j$. The construction of each $B_j$ corresponds to making a decision on which edges to include and which edges not to include. Then let
\begin{equation}
C_{F,2} = \{ e \in C_{F,1} \st e \in B_j \txt{ for some } j \in J\}.
\end{equation}
\begin{figure}[htb!]
	\begin{center}
	\includegraphics[width=3in]{images/analysis_block_collision}
	\end{center}
	\vspace{-20pt}
\caption{Example variable node indicating a collision.}
\label{f:collision}
\end{figure}
